---
title: "Intro to CriticalLineAlgo"
author: "Alejandro Sotolongo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6
)
```

```{r}
#devtools::install_github('alejandro-sotolongo/CriticalLineAlgo')
library(CriticalLineAlgo)
```

The critical line algorithm (CLA) was developed by Harry Markowitz to solve the corner portfolios of the efficiency frontier in his famous 1952 *Portfolio Selection* research. An open source version for Python exists and is outlined in the 2013 paper below. `CriticalLineAlgo` is a lightweight R implementation with minimal imports. The only outside package is `MASS` for general (Moore-Penrose) matrix inversion.

This package mostly follows the notation of [Bailey and Lopez de Prado 2013](https://papers.ssrn.com/sol3/papers.cfm?abstract_id=2197616) and [Niedermayer and Niedermayer 2006](https://papers.ssrn.com/sol3/papers.cfm?abstract_id=894842).

There are two main functions to interface with: `run_cla` and `calc_target_vol`. `run_cla` will solve for the corner portfolios and output a list with the stored output. `calc_target_vol` takes the list output from `run_cla` and finds the weights of an efficient portfolio for a specific level of volatility. 

In order to run the CLA a `mu_vec` of estimated returns and `cov_mat` of estimated covariance are needed. This introduction reads the `mu_vec` and `cov_mat` from csv files stored in Google Drive. The estimates are historical `mean` and `cov` estimates from 16 ETF time-series to represent different asset classes. The `mu_vec` and `cov_mat` are not predictions of the future. Instead they are used here to demonstrate how this package works.

```{r}
mu_data <- read.csv(sprintf('https://docs.google.com/uc?id=%s&export=download', 
                           '1vUPvKqJSR-3ZlWTWwf1r2kjguEDqmDM6'), 
                   row.names = NULL)
cov_data <- read.csv(sprintf('https://docs.google.com/uc?id=%s&export=download', 
                           '1g53v1D9mXvfcYKMsquHIj2CF7Lrn6-GX'),
                    row.names = NULL)
asset_class <- mu_data$ETF
mu_vec <- mu_data$mu
cov_mat <- as.matrix(cov_data[, 2:ncol(cov_data)])
mu_data
```

`run_cla` is the interface that runs the algorithm and returns the results. The `mu_vec` and `cov_mat` are passed in as the first two arguments.

```{r}
store <- run_cla(mu_vec, cov_mat)
```

The `$wgt_list` in the `store` output contains the weights of the corner portfolios. These weights can easily be combined and used to calculate the corner portfolios' expected return and volatility.

```{r}
corner_wgt <- do.call(cbind, store$wgt_list)
exp_ret <- apply(corner_wgt, 2, function(m, w) {t(w) %*% m}, m = mu_vec)
exp_vol <- apply(corner_wgt, 2, function(covar, w) {sqrt(t(w) %*% covar %*% w)},
                 covar = cov_mat)
plot(sqrt(diag(cov_mat)), mu_vec, col = 'grey', 
     ylab = 'Expected Return', xlab = 'Expected Volatility')
points(exp_vol, exp_ret, col = 'darkgreen', lwd = 2, type = 'b')
title('Efficiency Frontier Example')
mtext('Corner Portfolios in Green, Individual Assets in Grey')
rownames(corner_wgt) <- asset_class
colnames(corner_wgt) <- paste('Vol ', round(exp_vol * 100, 1))
round(corner_wgt * 100, 1)
```

`run_cla` supports adding lower and upperbound weight limits. This version of the CLA has a long-only constraint (i.e., the sum of the portfolio weights = 1). There are two ways of entering asset constraints. A single numeric value will be applied to each asset, e.g., entering `0.5` as the upper bound constraint will limit every asset to a maxium weight of 50%. A vector of weights that correspond to each asset in the `mu_vec` and `cov_mat` can be entered as input to allow for different constraints for each asset. The code below is an example of the latter: a constraint of 50% will be applied to all assets except HYG, AMJ, and GLD will be constrained at 10%.

```{r}
up_bound <- rep(0.5, length(mu_vec))
up_bound[asset_class %in% c('HYG', 'AMJ', 'GLD')] <- 0.1
store_constr <- run_cla(mu_vec, cov_mat, up_bound = up_bound)
corner_wgt <- do.call(cbind, store_constr$wgt_list)
exp_ret <- apply(corner_wgt, 2, function(m, w) {t(w) %*% m}, m = mu_vec)
exp_vol <- apply(corner_wgt, 2, function(covar, w) {sqrt(t(w) %*% covar %*% w)},
                 covar = cov_mat)
plot(sqrt(diag(cov_mat)), mu_vec, col = 'grey', 
     ylab = 'Expected Return', xlab = 'Expected Volatility')
points(exp_vol, exp_ret, col = 'darkgreen', lwd = 2, type = 'b')
title('Efficiency Frontier Example with Constraints')
mtext('Corner Portfolios in Green, Individual Assets in Grey')
rownames(corner_wgt) <- asset_class
colnames(corner_wgt) <- paste('Vol ', round(exp_vol * 100, 1))
round(corner_wgt * 100, 1)
```

`calc_target_vol` solves for the weights for an efficient portfolio with a specific volatility target. If the volatility target is not on the frontier a warning message will print and a value of `NA` will be returned.

```{r}
vol_10_wgt <- calc_target_vol(store_constr, cov_mat, 0.1)
rownames(vol_10_wgt) <- asset_class
round(vol_10_wgt * 100, 1)
sqrt(t(vol_10_wgt) %*% cov_mat %*% vol_10_wgt)
```

